Well-quasi-orderings on word languages

### 2024-10-17 src/amalgam.tex
moving from amalgam to infixes

\section{Amalgamation Systems and Decidability}
\label{infixes-amalgamation:sec}

When proving \cref{infix-finite-automata:thm}, we have leveraged a powerful
combinatorial argument on regular langugages stated in
\cref{pumping-periods:lem}. Our first remark is that this idea can
be generalized to the larger class of \intro{bounded languages}, i.e.,
languages $L \subseteq \Sigma^*$ such that there exists words $w_1, \dots, w_n$
satisfying $L \subseteq w_1^* \cdots w_n^*$.

@@

It turns out that there is a rather large family of systems for which pumping
arguments based on so-called \emph{minimal runs} exist: they are called
\emph{amalgamation systems} and were recently introduced by Anad, Schmitz,
Sch\"{u}tze, and Zetzsche \cite{ASZZ24}. Having done the heavy lifting on
\kl{bounded languages}, the rest of this section is mostly devoted to the
introduction of \kl{amalgamation systems} and collecting the necessary pumping
argument they enjoy. Using this meta-proof, we will generalize
\cref{infix-finite-automata:thm} to context-free grammars, languages recognized
by vector addition systems with state (VASS), and more. The goal of this
section is not to introduce all of these systems, or justify their usefulness,
but to state and prove the following theorem.

\begin{theorem}[label=infix-amalgamation:thm,restate=infix-amalgamation:thm]
    \proofref{infix-amalgamation:thm}
    Let $L \subseteq \Sigma^*$ be a language recognized by an 
    \kl{amalgamation system}.
    Then $L$ is well-quasi-ordered by the infix relation if and only if $L$ is
    a finite union of chains for the \kl{infix relation}.
\end{theorem}

@@

@@

\begin{lemma}
    \label{infix-amalgamation-bounded:lem}
    Let $L \subseteq \Sigma^*$ be a language recognized
    by an \kl{amalgamation system} that is \kl{well-quasi-ordered}
    for the \kl{infix relation}. Then $L$ is a \kl{bounded language}.
\end{lemma}
\begin{proof}
    Assume that $L$ is \kl{well-quasi-ordered} by the \kl{infix relation},
    and obtained by an \kl{amalgamation system}
    $(\Sigma, R, E, \canrun)$.

@@

    Because \kl{bounded languages} are stable under concatenation and finite
    unions, we conclude that $L$ itself must be a \kl{bounded language}.
\end{proof}

\begin{proofof}{infix-amalgamation:thm}
    We first apply \cref{infix-amalgamation-bounded:lem},
    and conclude with \cref{bounded-language:thm}.
\end{proofof}

For the next result, we impose a mild restriction by requiring the language class to be closed under rational transductions. This property holds for most common language classes, including all known \kl{amalgamation systems}.

\begin{theorem}
	\label{infix-wqo-is-emptiness:thm}
	Let $\mathcal{C}$ be a class of languages recognized by amalgamation systems and closed under rational transductions. Then the following are equivalent:

	\begin{enumerate}
		\item\label{wqo-infix-decidable} \kl[well-quasi-order]{Well-quasi-orderedness} of the \kl{infix relation} is decidable.
		\item\label{emptiness-decidable} Emptiness is decidable.
	\end{enumerate}
\end{theorem}

\begin{proof}
	\cref{emptiness-decidable} $\Rightarrow$ \cref{wqo-infix-decidable}. We aim to make the inclusion test of \cref{amalgamation-infix-wqo-inclusion:eqn} effective.

	Let $R = \bigcup_{x,y \in \Sigma^{\leq n}} \bigcup_{u \in \Sigma^{\leq m \times N_0}} \InfPeriodChain{x} u \InfPeriodChain{y} \cup \InfPeriodChain{x}u \cup u\InfPeriodChain{x}$. For any concrete values of the bounds $n$, $m$, and $N_0$, this language is regular. As $\mathcal{C}$ is closed under rational transductions, we can therefore reduce the inclusion to emptiness of $L \cap \overline{R}$. However, we need to find these bounds first.

@@

	To determine the value for $N_0$, we first computer maximal subsets $S \subseteq [n]$ such that $w_i$ for $i \in S$ are simultaneously unbounded. We do this by applying a transduction mapping $w_i$ to some fresh letter $a_i$ if $i \in S$ and $\varepsilon$ otherwise, and testing for simultaneous unboundedness \cite[Section 4.1]{ASZZ24}. Given a candidate bound $n_0$ for $N_0$, we construct the languages $L_{S,n_0} = w_1^{\circ_1}\cdots w_n^{\circ_n}$ where $\circ_i = *$ if $i \in S$ and $\circ_i = \leq n_0$ otherwise. We then test if $L \subseteq \bigcup_{S \text{ maximal}} L_{S,n_0}$. If yes, $n_0$ is our value for $N_0$, otherwise, we increase it and repeat the construction. As we chose maximal sets $S$, this procedure will eventually terminate.

	\cref{wqo-infix-decidable} $\Rightarrow$ \cref{emptiness-decidable}. We 
	consider the transduction $\mathcal T = \Sigma^* \times \{a, b\}$. Then 
	for any language $L \in \mathcal C$, $\mathcal TL$ is well-quasi-ordered 
	by infix if and only if $L$ is empty.
\end{proof}

>>>>

\section{Decision Prodecure and Amalgamation Systems}
\label{infixes-amalgamation:sec}

\AP
In this section, we are going to design an effective decision procedure for
\kl{well-quasi-ordering} by the \kl{infix relation}. To that end, the first
requirement is to fix a way to represent languages $L \subseteq \Sigma^*$.
Traditionally, one would use finite automata, context-free grammars, or for the
more adventurous, vector addition systems with states. However, our proof
technique will only require us to have a way to ``glue'' together runs of the
system to ``pump'' them and produce new runs: this is the usual pumping lemma
in automata theory, and Ogden's lemma for context-free grammars.
\todo{Cite the pumping lemma for VASS. Cite Ogden's lemma.}

\AP It turns out that there is a rather large family of systems for which
pumping arguments based on so-called \emph{minimal runs} exist: they are called
\emph{amalgamation systems} and were recently introduced by Anad, Schmitz,
Sch\"{u}tze, and Zetzsche \cite{ASZZ24}. Having done the heavy lifting on
\kl{bounded languages}, the rest of this section is mostly devoted to the
introduction of \kl{amalgamation systems} and collecting the necessary pumping
argument they enjoy. Using this meta-proof, we will generalize
\cref{infix-finite-automata:thm} to context-free grammars, languages recognized
by vector addition systems with state (VASS), and more.
\todo{unclear formulation}
The goal of this
section is not to introduce all of these systems, or justify their usefulness,
but to state and prove the following two theorems.

\begin{theorem}[label=infix-amalgamation:thm,restate=infix-amalgamation:thm]
    \proofref{infix-amalgamation:thm}
    Let $L \subseteq \Sigma^*$ be a language recognized by an 
    \kl{amalgamation system}.
    Then $L$ is well-quasi-ordered by the infix relation if and only if $L$ is
    \kl(language){bounded}.
\end{theorem}

\begin{theorem}[restated=infix-wqo-is-emptiness:thm,label=infix-wqo-is-emptiness:thm]
% 	\label{infix-wqo-is-emptiness:thm}
	Let $\mathcal{C}$ be a class of languages recognized by 
    \kl{amalgamation systems}
    and closed under
    \kl{rational transductions}. 
    Then the following are equivalent:
	\begin{enumerate}
        \item\label{wqo-infix-decidable} \kl[wqo]{Well-quasi-orderedness} of the \kl{infix relation} is decidable for languages in $\mathcal{C}$.
        \item\label{wqo-prefix-decidable} \kl[wqo]{Well-quasi-orderedness} of the \kl{prefix relation} is decidable for languages in $\mathcal{C}$.
        \item\label{emptiness-decidable} Emptiness is decidable for languages in $\mathcal{C}$.
	\end{enumerate}
\end{theorem}

@@

@@

\subsection{Amalgamation Systems, WQOs, and Bounded Languages}

@@

\begin{proofof}{infix-amalgamation:thm}
    Assume that $L$ is \kl{well-quasi-ordered} by the \kl{infix relation},
    and obtained by an \kl{amalgamation system}
    $(\Sigma, R, E, \canrun)$.

@@

    Because \kl{bounded languages} are stable under concatenation and finite
    unions, we conclude that $L$ itself must be a \kl{bounded language}.
\end{proofof}

One can leverage the same kind of reasoning to take the opposite direction, and
notice that the notions of being a \kl{bounded language}, a \kl{regular
language}, or being included in a finite union of products of \kl{chains} all
collapse for \kl{well-quasi-ordered} and \kl{downwards closed} languages for
the \kl{infix relation}.

\begin{lemma}
    \label{dwclosed-infixes-wqo:lem}
    Let $L \subseteq \Sigma^*$ be a \kl{downwards closed} language for the
    \kl{infix relation} that is \kl{well-quasi-ordered}. Then, the following
    are equivalent:
    {\renewcommand{\theenumi}{\roman{enumi}}
     \renewcommand{\labelenumi}{(\theenumi)}
    \begin{enumerate}
        \item\label{dwci-reg:item} $L$ is a \kl{regular language},
        \item\label{dwci-aml:item} $L$ is recognized by \emph{some} \kl{amalgamation system},
        \item\label{dwci-bod:item} $L$ is a \kl{bounded language},
        \item\label{dwci-uoc:item} $L$ is included in a finite union of products of \kl{chains}
    \end{enumerate}
    }
\end{lemma}
\begin{proof}
    It is clear that \cref{dwci-reg:item} $\Rightarrow$ \cref{dwci-aml:item}
    because regular languages are recognized by finite automata, and finite
    automata are a particular case of \kl{amalgamation systems}.
    The implication \cref{dwci-aml:item} $\Rightarrow$ \cref{dwci-bod:item}
    is the content of \cref{infix-amalgamation:thm}.
    The implication \cref{dwci-bod:item} $\Rightarrow$ \cref{dwci-uoc:item}
    is \cref{bounded-language:thm}.
    Finally, the implication \cref{dwci-uoc:item} $\Rightarrow$ \cref{dwci-reg:item}
    is simply because a \kl{downwards closed} language 
    that is a finite union of products of \kl{chains} is a regular language.
\end{proof}

\subsection{Effective Decision Procedures}
\label{infixes-amalgamation-effective:subsec}

For the next result, we impose a mild restriction by requiring the language
class to be closed under \kl{rational transductions}. This property holds for
most common language classes, including all known \kl{amalgamation systems}.

\AP A class of languages $\mathcal{C}$ is \intro{recognized by amalgamation
systems} when, for every language $L \in \mathcal{C}$, there exists an
\kl{amalgamation system} $(\Sigma, R, E, \canrun)$ such that $L$ is the
language of this system.

\AP A \intro{rational transduction} is a function $f \colon \Sigma^* \to
\Gamma^*$ that can be defined by a finite automaton with output, or
equivalently by a finite set of rules of the form $u \to v$ where $u \in
\Sigma^*$ and $v \in \Gamma^*$. We write $f(L) = \setof{f(w)}{w \in L}$ for the
image of a language $L$ by a rational transduction $f$. A class of languages
$\mathcal{C}$ is \intro{closed under rational transductions} when, for every
language $L \in \mathcal{C}$, and every rational transduction $f \colon
\Sigma^* \to \Gamma^*$, the language $f(L)$ is in $\mathcal{C}$.\todo{check these definitions ...}

\begin{proofof}{infix-wqo-is-emptiness:thm}
	\cref{emptiness-decidable} $\Rightarrow$ \cref{wqo-infix-decidable}. We aim to make the inclusion test of \cref{bounded-language:eq} effective.

    Let $R = \bigcup_{x,y \in \Sigma^{\leq n}} \bigcup_{u \in \Sigma^{\leq m
    \times N_0}} \InfPeriodChain{x} u \InfPeriodChain{y} \cup
    \InfPeriodChain{x}u \cup u\InfPeriodChain{x}$. For any concrete values of
    the bounds $n$, $m$, and $N_0$, this language is regular. As $\mathcal{C}$
    is \kl{closed under rational transductions}, we can therefore reduce the
    inclusion to emptiness of $L \cap \overline{R}$.\todo{explain that $L
    \mapsto L \cap R$ is a rational  transduction?} However, we need to find
    these bounds first.

@@

    To determine the value for $N_0$, we first computer maximal subsets $S
    \subseteq [n]$ such that $w_i$ for $i \in S$ are simultaneously unbounded.\todo{define this}
    We do this by applying a transduction mapping $w_i$ to some fresh letter
    $a_i$ if $i \in S$ and $\varepsilon$ otherwise, and testing for
    simultaneous unboundedness \cite[Section 4.1]{ASZZ24}. Given a candidate
    bound $n_0$ for $N_0$, we construct the languages $L_{S,n_0} =
    w_1^{\circ_1}\cdots w_n^{\circ_n}$ where $\circ_i = *$ if $i \in S$ and
    $\circ_i = \leq n_0$ otherwise. We then test if $L \subseteq \bigcup_{S
    \text{ maximal}} L_{S,n_0}$. If yes, $n_0$ is our value for $N_0$,
    otherwise, we increase it and repeat the construction. As we chose maximal
    sets $S$, this procedure will eventually terminate.

    \cref{wqo-infix-decidable} $\Rightarrow$ \cref{wqo-prefix-decidable}. We
    just consider the transduction

    \cref{wqo-prefix-decidable} $\Rightarrow$
    \cref{emptiness-decidable}. We consider the transduction $\mathcal T
    = \Sigma^* \times \{a, b\}$.\todo{Define this transduction explicitly} Then
    for any language $L \in \mathcal C$, $\mathcal TL$ is well-quasi-ordered by
    infix if and only if $L$ is empty.
\end{proofof}
